---
title: "Research Proposal"
author: "Belicia Rodriguez"
date: "10/11/2019"
output:
  html_document:
    code_folding: hide
    toc: yes
    theme: readable
---

```{r, include=F}
library(tidyverse)
library(wooldridge)
library(stargazer)
library(knitr)
```

*Download dataset*
```{r}
data(bwght2, package='wooldridge')
```

**Section I**

*What is your research question? Why is it interesting? What is the typical understanding of the phenomena you are studying, and how will your paper add to or change this understanding?*

I want to know whether prenatal behaviors such as prenatal visits, smoking cigarettes, and drinking alcohol positively or negatively affect a baby's birth weight.

If there is either a positive or negative relationship, then the question would be satisfied in showing that prenatal behaviors do affect a baby's birth weight. Generally, one would believe that prenatal visits would positively affect a baby's birth weight while smoking cigarettes and drinking alcohol would negatively affect a baby's birth weight. This study, however, will show whether the presumed-positive effect of prenatal visits will be greater than the presume-negative effect of smoking cigarettes and drinking alcohol. Also, this study will look at the percentage increase or decrease in birth weight instead of the birth weight in grams in order to observe the positive or negative effects as a percentage change.

The American Pregnancy Association says that smoking during pregnancy causes "a greater chance of having a baby that weighs too little." (1) They also say that when a mother drinks "excessively" during pregnancy there can be "complications which can occur during pregnancy," and that "these risks may not be associated as strongly with occassional drinking." (2) This information suggest that the affect of drinking, on average, one or two drinks per week during pregnancy may not have a big effect on baby weight in our model. Also, a study on whether the relationship between prenatal care and birth weight vary by oral clefts at the National Institutes of Health shows that for babies born in the USA and unaffected by oral clefts had an increased birth weight, on average, of 16 grams per prenatal visit. Therefore, these studies show the possible negative affects of smoking and drinking and possible positive effect of prenatal visits, but they do not show whether smoking, drinking, or prenatal visits affect a baby's weight more than the others.

*Graphs and/or tables are a must here.*

```{r, message=FALSE, echo=FALSE}
# scatter plot of number of prenatal vists vs log of baby's weight
ggplot(bwght2, aes(npvis, lbwght)) +
  geom_point() +
  geom_smooth(method="lm", se= FALSE, linetype = "dashed",aes(col="Reg. Line"))+
  ggtitle("Number of Prenatal Vists vs Log of Baby's Weight")

# scatter plot of average cigarettes per day vs log of baby's weight
ggplot(bwght2, aes(cigs, lbwght)) +
  geom_point() +
  geom_smooth(method="lm", se= FALSE, linetype = "dashed",aes(col="Reg. Line"))+
  ggtitle("Average Number of Cigarettes vs Log of Baby's Weight")

# scatter plot of average number of drinks per week vs log of baby's weight
ggplot(bwght2, aes(drink, lbwght)) +
  geom_point() +
  geom_smooth(method="lm", se= FALSE, linetype = "dashed",aes(col="Reg. Line"))+
  ggtitle("Average Number of Drinks vs Log of Baby's Weight")

```


**Section II**

*What data are you going to use? Discuss in further detail the population of study and any important characteristics of the data.*

My research question will look at the variables lbwght (log of birth weight), npvis (total number of prenatal visits), cigs (avgerage cigarettes per day), and drink (average drinks per week). The population of study is newborn babies, and the newborns are described by the information of their parents. Each observation has information about their births such as apgar score and birth weight, but most of the information this dataset contains on the newborn is characteristics of the mother and father such as race, years of education, and the mother's prenatal habits (i.e smoking, drinking, prenatal visits).

*Graphs and tables of descriptive statistics on the data usually accompany this section.*

When assembling these descriptive statistics, I decided to solely focus on the mother's information and not the father's because this research proposal focuses on the mother's prenatal habits and how it affects the baby. Therefore, I am more interested in understanding the demographic of the mothers in this study.


```{r, results="markuo"}
# look at the structure of the data
str(bwght2)
```

Proportion of categorical variables that are not in model but help understand demographic of the population
```{r, results="markup"}
# proportion of different categorical variables
prop.table(table(factor(bwght2$lbw, labels=c("not low", "low bweight"))))
prop.table(table(factor(bwght2$male, labels=c("female baby", "male baby"))))
prop.table(table(factor(bwght2$mwhte, labels=c("mother not white", "mother white"))))
prop.table(table(factor(bwght2$mblck, labels=c("mother not black", "mother black"))))
prop.table(table(factor(bwght2$moth, labels=c("mother not other", "mother other race"))))
```

```{r, results = 'asis'}
bwght2 %>% select(lbw, male, mwhte, mblck, mage, meduc) %>% summarize_all(mean, na.rm=TRUE)%>% t() %>% data.frame()%>%setNames(., c("Average")) %>% kable()
```


Average mother's age
```{r}
mean(bwght2$mage)
```

Average's mothers years of education
```{r}
# average mother's years of education
mean(bwght2$meduc, na.rm=T)
```

```{r, results="asis", tidy = T}
# Create a table that displays the number of samples, mean, standard deviation, minimum and maximum of all the variables in the prenatal habits variable data frame; set the type to html in order to print in html file

stargazer(as.data.frame(bwght2 %>% select(lbwght, npvis, npvissq, cigs, drink)), title = "Statistical Analysis of Prenatal Habits Variables", summary.stat=c("n", "mean", "sd", "min", "max"), type="html")
```


**Section III**

*Motivate and present the economic model. Discuss why what you are about to do makes sense.*

Economic model: birth_weight = f(prenatal_vists, avg_cigarettes, avg_drink)

My economic model shows a relationship between number of prenatal visits, average number cigarettes per day, and average number of drinks per week that will affect the birth weight. Prenatal visits, average number of cigarettes per day, and average number of drinks per week are numbers that reflect the behaviors of a mother throughout the months of her pregnancy, and many would speculate that a mother's habits thoughout her pregnancy will affect the health of her baby, which in this case will evaluated by assessing whether these variables cause a percentage change in the birth weight. Therefore, choosing these three variables seems like a natural choice to see whether going to prenatal visits and monitoring a mother's progress will increase the baby's weight and whether smoking and drinking at all during pregnancy will decrease the baby's weight

*From the general economic model (or intuition), you should specify and discuss an econometric model. What is your chosen specification, and why? Show your equations here.*

Econometric model: log(bwght) = $\beta_0$ + $\beta_1$npvis + $\beta_2$npvissq + $\beta_3$cigs + $\beta_4$drink + u

The econometric model is a multiple linear regression with four explanatory variables where the npvis (number of prenatal visits) represents a quadratic relationship between number of prenatal visits and the percentage change of birth weight. Therefore, depending on the which parameters ($\beta_0$, $\beta_1$, $\beta_2$, $\beta_3$, and $\beta_4$) have larger or smaller values, the linear regression may have an overall more linear shape or more quadratic shape, depending on whether the absolute value of $\beta_1$ and $\beta_2$ is higher or lower than the absolute value of $\beta_3$ and $\beta_4$. 


**Bibliography**

1. “Smoking During Pregnancy.” American Pregnancy Association, 20 Sept. 2019, [americanpregnancy.org](americanpregnancy.org/pregnancy-health/smoking-during-pregnancy/#targetText=You%20have%20a%20greater%20chance,disabilities%2C%20and%20physical%20growth%20problems.)

2. Nyarko, Kwame A, et al. “Does the Relationship between Prenatal Care and Birth Weight Vary by Oral Clefts? Evidence Using South American and United States Samples.” The Journal of Pediatrics, U.S. National Library of Medicine, Jan. 2013, [www.ncbi.nlm.nih.gov](www.ncbi.nlm.nih.gov/pmc/articles/PMC3485451/.)


**Other Important Information**

*Variables in dataset*

Name |  Description
-----------|--------
 mage   | mother's age, years
 meduc  | mother's educ, years
 monpre  | mother prenatal care began
 npvis | total number of prenatal visits
 fage | father's age, years
 feduc  | father's educ, years
 bwght | birth weight, grams
 omaps | one minute apgar score
 fmaps | five minute apgar score
 cigs | avg cigarettes per day
 drink | avg drinks per week
 lbw | =1 if bwght <= 2000
 vlbw | =1 if bwght <= 1500
 male | =1 if baby male
 mwhte | =1 if mother white
 mblck | =1 if mother is black
 moth | =1 if mother is other
 fwhte | =1 if father white
 fblck | =1 if father black
 foth | =1 if father is other
 lbwght | log(bwght)
 magesp | mage^2
 npvissq | npvis^2

definitions: apgar score

- total score of 1 to 10

- higher the score, the better the baby is doing after birth

- score of 7,8, or 9 is normal and a sign the newborn is in good health

one minute apgar score

- performed on baby a minute after birth

- determines how well a baby tolerated birthing process

five minute apgar score

- says how well baby is doing outside mother's womb


FOR FINAL PROJECT (not submitted)

# Introduction and motivation:

What is your research question? Why is it interesting? What is the typical understanding of the phenomena you are studying, what is the related literature and how will your paper add to or change this understanding? Graphs and/or tables are a must here. Provide a broad idea of the data from Wooldridge and external data to support your claims and hypothesis. External data can be form other studies, blogs, surveys, tables. etc. 

# The data set 

1. Briefly describe your data source and data. Why do you think this data is adequate to answer your research question? Do the required manipulations so that your data have self explanatory names and if not, describe those variables shortly. 

2. Discuss in further detail the population of study and any important characteristics of the data. (Graphs and tables of descriptive statistics on the data usually accompany this section.)

Motivate your regression analysis and research hypothesis with graphs and descriptive stats from your data. Explain some of the important descriptive statistics of your data set. Remember to use stargazer to do a summary of your data set and add other tables if needed. Use ggplot to do graphs that describe important characteristics or features of your data 

```{r }
# use stargazer to do a table of the summary statistics of your dataset 
```

#  Methodology/Empirical Model:
Here you motivate and present the economic model (regression equations). Discuss why what you are about to do makes sense. From the general economic model (or intuition), you should specify and discuss an econometric model. What is your chosen specification, and why? Show your equations here.

Write down the regression equation and/or equations, see the math below. Show as many equations as regressions you will estimate, the more the merrier as long as they make sense. Do not run regression if you don't have a clear idea about why you are doing that. 
For example: I will estimate the following equation 

Basic Model$$ y = \beta_0+\beta_1*x_1 + \beta_2*x_2+\beta_3*x_3 +u$$
Model with bla $$ y = \beta_0+\beta_1*x_1 + \beta_2*x_2+\beta_3*x_3 + \beta_4*x_4+u$$
Model with bla $$ y = \beta_0+\beta_1*x_1 + \beta_2*x_2+\beta_3*x_3 + \beta_4*x_4+u$$
Model interacting bla and bla and dummy of bla $$ y = \beta_0+\beta_1*x_1 + \beta_2*x_2+\beta_3*(x_3*dummy1) + \beta_4*x_4+u$$


# Results

Present results of econometric estimations (in a table) interpret and discuss. Typically present between 3 and 15 different specifications depending on different robustness checks you want to make. The latter part of this section should also include a detailed discussion of many of the course topics including OVB, heteroskedasticity, tests of multiple and joint hypotheses, etc. and what you did pertinent to these topics.

Remember when using startgazer to put toguether the results of your regressions. See in the chunk of code below that the option results='asis' is added to the options of the code. 

```{r, results='asis'}

```

-	Run at least 3 regressions ( As I mentioned I expect that you have more than two variables so that you are able to do multiple regression analysis and not only simple). 
-	Talk about the results and the statistical significance of parameters. 
-	If you are using qualitative or categorical data do the adequate manipulations in your regression to add these. Experiment changing the reference group. This will make sense after you see this topic in class. If you have categorical variables, which ideally you do, you should do this. 

Put all your regressions results in the same stargazer table if possible, as long as they relate. Do not repeat the results twice with and without stargazer. 

Do some inference test and describe your results 

-	t test (in some case you will need it if you have a hypothesis differnt that the default in the lm command)
-	F test (definitelly perfom an F test for your results, if you use the one in the lm command explain)

# Conclusion
Summarize the approach of your study, what you tried to explain, and the highlights of your results. Restate any caveats to or shortcomings of your approach (every study has some), and what could be done in future work to overcome these issues.

# References
(separate from the body of text) Works cited/ references. Full references to the works you cited in the motivation sections. At least 3 references.








Choose only the variables that are part of your regression analysis. And create the data set with those, do not print your all your dataset into the Rmarkdown document, no need to presnet the strut\cture only relevent descriptive stats tables. 

If you have issues with your data structure it might be because it is not in the right format. I will upload a handout about using the package tidyr that allows you to improve the change the format of your dataset. Also check the tips of the week section on canvas there are some usefull tips there. Unlikely if using wooldridge data. 










